Almost orthogonal vectors

### 2014-12-11 project3/paper/paper.tex
More changes to paper

\section{Introduction}
Consider a collection of $N$ unit vectors $v_1, \ldots, v_N$ in $\mathbb R^n$. Define $\epsilon$ as
\[\epsilon=\max_{i\neq j}|v_i\cdot v_j|^2.\]
What is the minimal attainable $\epsilon$?

For $n=N$, it is clear that the optimal value of $\epsilon$ is zero, obtained by letting $v_1,\ldots, v_N$ be any orthonormal basis. And for $N>n$, we must have $\epsilon>0$ as any collection of $N$ vectors is linearly dependent, while any collection of orthonormal vectors is linearly independent.

@@

Again without loss of generality, we can assume that the vectors $v_1, v_2, \ldots, v_N$ are on or above the $x$-axis, and that $v_1 = [1,0]^T$ -- if any vector $v_i$ were not above the $x$-axis, then we could just consider $-v_i$ (the reflection of $v_i$ about the $x$-axis) instead.

@@

Recall the Cauchy-Schwarz for $\mathbb R^n$: given $(x_1,\ldots, x_n)$ and $(y_1,\ldots, y_n)\in \mathbb R^n$, we have
\[\left(\sum_{i=1}^nx_iy_i\right)^2\leq \left(\sum_{i=1}^n x_i^2\right)\left(\sum_{i=1}^ny_i^2\right).\]
In our case, let the $x_i$ equal $\lambda_i$, and let each $y_i$ equal 1 when $\lambda_i$ is nonzero and zero when $\lambda_i$ is zero. We get
\[n^2=\left(\sum_{i=1}^n \lambda_i\right)^2\leq \left(\sum_{i=1}^n \lambda_i^2\right) k\]
so that
\[\sum_{i=1}^n \lambda_i^2\geq\frac{n^2}{k}\geq \frac{n^2}{d}.\]
In fact, the sum $\sum_{i=1}^n \lambda_i^2$ is equal to the trace of $A^TA$, which we can compute explicitly to be $\sum_{i,j}a_{i,j}^2$. Hence we have that
\[\sum_{i,j}a_{i,j}^2\geq \frac{n^2}{d}.\]
But because $|a_{i,j}|\leq \sqrt\epsilon$ for $i\neq j$, we can bound the left hand side:
\[n+n(n-1)\epsilon\geq \left(\sum_{i=1}^j a_{i,i}^2\right)+\left(\sum_{i\neq j}a_{i,j}^2\right)=\sum_{i,j}a_{i,j}^2.\]
We obtain, therefore, that
\[n+n(n-1)\epsilon\geq \frac{n^2}{d},\]
and it clearly follows that
\[\epsilon\geq \frac{n-d}{d(n-1)}.\]
\end{proof}
Using this lemma, we can now prove our lower bound on $\epsilon$.
\begin{theorem}
\label{lower_bound}
For any choice of $n$ and $N$, the minimum value of $\epsilon$ satisfies
\[\epsilon\geq \frac{N-n}{n(N-1)}.\]
\end{theorem}
\begin{proof}
Pick any collection of $N$ unit vectors $v_1,\ldots, v_N$ in $\mathbb R^n$, and define $\epsilon$ as usual. Recall the definitions of $A$ and $G$:
\[A=\left(\begin{matrix}
| & | & \cdots & | & |\\
v_1 & v_2 & \cdots & v_{N-1} & v_N\\
| & | & \cdots & | & | \end{matrix}\right).\] 
\[G(v_1,\ldots, v_N)=A^TA=\left(\begin{matrix}
v_1\cdot v_1 & v_1\cdot v_2 & \ldots & v_1\cdot v_n\\
v_2\cdot v_2 & v_2\cdot v_2 & \ldots & v_2\cdot v_n\\
\vdots & \vdots & \ddots & \vdots\\
v_n\cdot v_1 & v_n\cdot v_2 & \ldots & v_n\cdot v_n\end{matrix}\right).\]
Let $d$ be the rank of $G$. Since $A$ has rank at most $n$, so does $A^T$; since the image space of $G$ must be contained in the image space of $A^T$, we have that $d\leq n$. Note that $G$ is real and symmetric, with $1$s along the diagonal and all other entries with absolute value bounded by $\sqrt \epsilon$. We can apply the lemma to conclude
\[\epsilon\geq \frac{N-d}{d(N-1)}.\]
Since $d\leq n$, we have as desired
\[\epsilon\geq \frac{N-n}{n(N-1)}.\]
\end{proof}
We present now a table of the bounds the theorem gives us for varying $n$ and $N$:
\begin{table}[h]
   \centering
    \begin{tabular}{ | c | c | c | c |}
    \hline
      & $n=2$ & $n=3$ & $n=4$ \\ \hline
    $N=2$ & $0$ & $-$ & $-$ \\ \hline
    $N=3$ & $1/4$ & $0$ & $-$ \\ \hline
    $N=4$ & $1/3$ & $1/9$ & $0$ \\ \hline
    $N=5$ & $3/8$ & $1/6$ & $1/16$ \\ \hline
    $N=6$ & $2/5$ & $1/5$ & $1/10$ \\ \hline
    $N=8$ &  $3/7$ & $5/21$ & $1/7$ \\ \hline
    $N=12$ & $5/11$ & $3/11$ & $2/11$ \\
    \hline
    \end{tabular}
    \caption {Lower bounds on $\epsilon$ for different values of $(n, N)$, given by Theorem \ref{lower_bound}.}
    \end{table}

Note that the bounds for $n=2$ are not tight, as we proved in Section 2 that the minimum $\epsilon$ was given by $\cos^2(\pi/N)$. We shall see, however, that the bounds for $\epsilon$ are in fact tight in the cases where $N=n+1$ and where $n=3, N=6$. One direction for further research might be checking whether the bounds provided by Theorem \ref{lower_bound} are indeed attainable for other combinations of $n$ and $N$.

@@

\begin{theorem}
If $N=n+1$, then the minimum value of $\epsilon$ for $i\neq j$ is $\frac{1}{n^2}$. This $\epsilon$ is achieved when the vectors are arranged in an $n$ dimensional simplex centered at the origin.
\label{simplex}
\end{theorem}

\begin{proof}
Let $\epsilon$ be as before for a given set of unit vectors $v_1, v_2, \ldots, v_{n+1}$. We show that the origin-centered simplex configuration of the vectors $v_1,\cdots v_{n+1}$ produces a value $\epsilon=\frac{1}{n^2}$.

@@

Though this approach seems simple, it produces  results which seem to agree with the $\epsilon$ we derived for $(2, N)$ in Section 2, and with the $\epsilon$ we derived for $(3, 6)$ and $(n, n+1)$ in Section 4. This gives us some confidence in the accuracy of our algorithm. However, it must be noted that the data do not agree with the bounds produced by Theorem \ref{lower_bound}. There are multiple interpretations of this fact. First, it is possible that the lower bounds produced by Theorem \ref{lower_bound} are not always tight; they in fact are not for the case $(2, N)$. Alternatively, it is possible that our code simply becomes inaccurate for larger $n$ and $N$. This could be a fruitful direction for further research.

@@

Our results on this section concern the case when $N=n+1$. We shall see, in fact, that the optimal value of $\delta$ is given by $-\frac{1}{n}$, we introduce the parallelogram law.

@@

First we give the definition of the norm on vectors in $\R^n$ in terms of the dot product:

@@

\subsection{Variations: Minimizing the dot product}
Given the above bound, we proceed to state and prove the following theorem.

>>>>

\section{Introduction}
Consider a collection of $N$ unit vectors $v_1, \ldots, v_N$ in $\mathbb R^n$. Define $\epsilon$ as
\[\epsilon=\max_{i\neq j}|v_i\cdot v_j|^2.\]
In this paper, we try to address the question of what the minimal attainable $\epsilon$ is.

For $n=N$, it is clear that the optimal value of $\epsilon$ is zero, obtained by letting $v_1,\ldots, v_N$ be any orthonormal basis. And for $N>n$, we must have $\epsilon>0$ as any collection of $N$ vectors in $n$-space is linearly dependent, while any collection of orthonormal vectors is linearly independent.

@@

Again without loss of generality, we can assume that the vectors $v_1, v_2, \ldots, v_N$ are on or above the $x$-axis, and that $v_1 = [1,0]^T$ -- if any vector $v_i$ were not above the $x$-axis, then we could just consider $-v_i$ (the reflection of $v_i$ about the origin) instead.

@@

Recall the Cauchy-Schwarz inequality for $\mathbb R^n$: given $(x_1,\ldots, x_n)$ and $(y_1,\ldots, y_n)\in \mathbb R^n$, we have
\[\left(\sum_{i=1}^nx_iy_i\right)^2\leq \left(\sum_{i=1}^n x_i^2\right)\left(\sum_{i=1}^ny_i^2\right).\]
In our case, let the $x_i$ equal $\lambda_i$, and let each $y_i$ equal 1 when $\lambda_i$ is nonzero and zero when $\lambda_i$ is zero. We get
\[n^2=\left(\sum_{i=1}^n \lambda_i\right)^2\leq \left(\sum_{i=1}^n \lambda_i^2\right) k\]
so that
\[\sum_{i=1}^n \lambda_i^2\geq\frac{n^2}{k}\geq \frac{n^2}{d}.\]
In fact, the sum $\sum_{i=1}^n \lambda_i^2$ is equal to the trace of $A^TA$, which we can compute explicitly to be $\sum_{i,j}a_{i,j}^2$. Hence we have that
\[\sum_{i,j}a_{i,j}^2\geq \frac{n^2}{d}.\]
But because $|a_{i,j}|\leq \sqrt\epsilon$ for $i\neq j$, we can bound the left hand side:
\[n+n(n-1)\epsilon\geq \left(\sum_{i=1}^j a_{i,i}^2\right)+\left(\sum_{i\neq j}a_{i,j}^2\right)=\sum_{i,j}a_{i,j}^2.\]
We obtain, therefore, that
\[n+n(n-1)\epsilon\geq \frac{n^2}{d},\]
and it clearly follows that
\[\epsilon\geq \frac{n-d}{d(n-1)}.\]
\end{proof}
Using this lemma, we can now prove our lower bound on $\epsilon$.
\begin{theorem}
\label{lower_bound}
For any choice of $n$ and $N$, the minimum value of $\epsilon$ satisfies
\[\epsilon\geq \frac{N-n}{n(N-1)}.\]
\end{theorem}
\begin{proof}
Pick any collection of $N$ unit vectors $v_1,\ldots, v_N$ in $\mathbb R^n$, and define $\epsilon$ as usual. Recall the definitions of $A$ and $G$:
\[A=\left(\begin{matrix}
| & | & \cdots & | & |\\
v_1 & v_2 & \cdots & v_{N-1} & v_N\\
| & | & \cdots & | & | \end{matrix}\right).\] 
\[G(v_1,\ldots, v_N)=A^TA=\left(\begin{matrix}
v_1\cdot v_1 & v_1\cdot v_2 & \ldots & v_1\cdot v_n\\
v_2\cdot v_2 & v_2\cdot v_2 & \ldots & v_2\cdot v_n\\
\vdots & \vdots & \ddots & \vdots\\
v_n\cdot v_1 & v_n\cdot v_2 & \ldots & v_n\cdot v_n\end{matrix}\right).\]
Let $d$ be the rank of $G$. Since $A$ has rank at most $n$, so does $A^T$; since the image space of $G$ must be contained in the image space of $A^T$, we have that $d\leq n$. Note that $G$ is real and symmetric, with $1$s along the diagonal and all other entries with absolute value bounded by $\sqrt \epsilon$. We can apply the lemma to conclude
\[\epsilon\geq \frac{N-d}{d(N-1)}.\]
Since $d\leq n$, we have as desired
\[\epsilon\geq \frac{N-n}{n(N-1)}.\]
\end{proof}
We present now a table of the bounds the theorem gives us for varying $n$ and $N$:
\begin{table}[h]
   \centering
    \begin{tabular}{ | c | c | c | c |}
    \hline
      & $n=2$ & $n=3$ & $n=4$ \\ \hline
    $N=2$ & $0$ & $-$ & $-$ \\ \hline
    $N=3$ & $1/4$ & $0$ & $-$ \\ \hline
    $N=4$ & $1/3$ & $1/9$ & $0$ \\ \hline
    $N=5$ & $3/8$ & $1/6$ & $1/16$ \\ \hline
    $N=6$ & $2/5$ & $1/5$ & $1/10$ \\ \hline
    $N=8$ &  $3/7$ & $5/21$ & $1/7$ \\ \hline
    $N=12$ & $5/11$ & $3/11$ & $2/11$ \\
    \hline
    \end{tabular}
    \caption {Lower bounds on $\epsilon$ for different values of $(n, N)$, given by Theorem \ref{lower_bound}.}
    \end{table}

Note that the bounds for $n=2$ are not tight, as we proved in Section 2 that the minimum $\epsilon$ is given by $\cos^2(\pi/N)$ when $n=2$. We shall see, however, that the bounds for $\epsilon$ are in fact tight in the cases where $N=n+1$ and where $n=3, N=6$. One direction for further research might be checking whether the bounds provided by Theorem \ref{lower_bound} are indeed attainable for other combinations of $n$ and $N$.

@@

\begin{theorem}
If $N=n+1$, then the minimum value of $\epsilon$ for $\frac{1}{n^2}$. This $\epsilon$ is achieved when the vectors are arranged in an $n$ dimensional simplex centered at the origin.
\label{simplex}
\end{theorem}

\begin{proof}
Let $\epsilon$ be defined as before for a given set of unit vectors $v_1, v_2, \ldots, v_{n+1}$. We show that the origin-centered simplex configuration of the vectors $v_1,\cdots v_{n+1}$ produces a value $\epsilon=\frac{1}{n^2}$.

@@

Though this approach seems simple, it produces  results which seem to agree with the $\epsilon$ we derived for $(2, N)$ in Section 2, and with the $\epsilon$ we derived for $(3, 6)$ and $(n, n+1)$ in Section 4. This gives us some confidence in the accuracy of our algorithm. However, it must be noted that the data does not agree with the bounds produced by Theorem \ref{lower_bound}. There are multiple interpretations for this fact. First, it is possible that the lower bounds produced by Theorem \ref{lower_bound} are not always tight -- this is easy to see for the case of $(2, N)$. Alternatively, it is possible that our code simply becomes inaccurate for larger $n$ and $N$. This could be a fruitful direction for further research.

@@

Our results on this section concern the case when $N=n+1$. We shall see, in fact, that the optimal value of $\delta$ is given by $-\frac{1}{n}$. To begin, we introduce the paralleogram law.

@@

First we give the definition of the norm of a sum of vectors in $\R^n$ in terms of the dot product:

@@

\subsection{Variation: Minimizing the dot product}
Given the above bound, we proceed to state and prove the following theorem.
